---
title: "scDesignPop Quickstart"
author: 
  - name: Chris Dong
    affiliation:
    - Department of Statistics and Data Science, University of California, Los Angeles
    email: cycd@g.ucla.edu
  - name: Yihui Cen
    affiliation:
    - Department of Computational Medicine, University of California, Los Angeles
    email: yihuicen@g.ucla.edu
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('scDesignPop')`"
vignette: >
  %\VignetteIndexEntry{scDesignPop-quickstart-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---
```{css, echo=FALSE}
pre {
  white-space: pre !important;
  overflow-x: scroll !important;
}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    message = FALSE,
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
tools::R_user_dir("scDesignPop", which="cache")
```

## Introduction
scDesignPop is a simulator for population-scale single-cell RNA-sequencing (scRNA-seq) data. For more information, please check the Articles on our website: (https://chrisycd.github.io/scDesignPop/docs/index.html).

## Library and data preparation

Here, we use an example SingleCellExperiment object `example_sce` with $982$ genes and $7998$ cells and an example eQTL genotype dataframe `example_eqtlgeno` to demonstrate the main tutorial. These two objects contains the gene expression and SNP genotypes of $40$ anonymized individuals while the eQTL genotype dataframe provides $2792$ putative cell-type-specific eQTLs.

```{r, message=FALSE, warning=FALSE}
library(scDesignPop)
library(SingleCellExperiment)
library(SummarizedExperiment)
library(ggplot2)
theme_set(theme_bw())

data("example_sce")
data("example_eqtlgeno")

head(colData(example_sce))
```

## Modeling and simulation

### Step 1: construct a data list

To run scDesignPop, a list of data is required as input. This is done using the `constructDataPop` function. A `SingleCellExperiment` object and an `eqtlgeno` dataframe are the two main inputs needed. The `eqtlgeno` dataframe consists of eQTL annotations (it must have cell state, gene, SNP, chromosome, and position columns at a minimum), and genotypes across individuals (columns) for every SNP (rows). The structure of an example `eqtlgeno` dataframe is also given below.

```{r}
head(example_eqtlgeno)
```

To extract the information from `example_eqtlgeno` and the metadata from `example_sce`, users can use `celltype_colname`, `feature_colname`, `snp_colname`, `loc_colname`, `chrom_colname` and `indiv_colname` to specify the correct column names in the `example_eqtlgeno` and `example_sce` during the pipeline.

```{r}
data_list <- constructDataPop(
    sce = example_sce,
    eqtlgeno_df = example_eqtlgeno,
    new_covariate = as.data.frame(colData(example_sce)),
    copula_variable = "cell_type",
    slot_name = "counts",
    snp_mode = "single",
    celltype_colname = "cell_type",
    feature_colname = "gene_id",
    snp_colname = "snp_id",
    loc_colname = "POS",
    chrom_colname = "CHR",
    indiv_colname = "indiv"
    )
```

### Step 2: fit marginal model

Next, a marginal model is specified to fit each gene using the
`fitMarginalPop` function.  
Here we use a Negative Binominal as the parametric model using `"nb"`.

```{r}
marginal_list <- fitMarginalPop(
    data_list = data_list,
    mean_formula = "(1|indiv) + cell_type",
    model_family = "nb",
    interact_colnames = "cell_type",
    parallelization = "parallel",
    n_threads = 20L,
    loc_colname = "POS",
    snp_colname = "snp_id",
    celltype_colname = "cell_type",
    indiv_colname = "indiv"
    )
```

### Step 3: fit a Gaussian copula

The third step is to fit a Gaussian copula using the `fitCopulaPop`
function.

```{r}
set.seed(123, kind = "L'Ecuyer-CMRG")

copula_fit <- fitCopulaPop(
    sce = example_sce,
    assay_use = "counts",
    input_data = data_list[["covariate"]],
    marginal_list = marginal_list,
    family_use = "nb",
    copula = "gaussian",
    n_cores = 2L,
    parallelization = "parallel"
    )

RNGkind("Mersenne-Twister")  # reset
```

### Step 4: extract parameters

The fourth step is to compute the mean, sigma, and zero probability
parameters using the `extractParaPop` function.

```{r}
para_new <- extractParaPop(
    sce = example_sce,
    assay_use = "counts",
    marginal_list = marginal_list,
    n_cores = 2L,
    family_use = "nb",
    indiv_colname = "indiv",
    new_covariate = data_list[["new_covariate"]],
    new_eqtl_geno_list = data_list[["eqtl_geno_list"]],
    data = data_list[["covariate"]],
    parallelization = "parallel"
    )
```

### Step 5: simulate counts

The fifth step is to simulate counts using the `simuNewPop` function.

```{r}
set.seed(123, kind = "L'Ecuyer-CMRG")

newcount_mat <- simuNewPop(
    sce = example_sce,
    mean_mat = para_new[["mean_mat"]],
    sigma_mat = para_new[["sigma_mat"]],
    zero_mat = para_new[["zero_mat"]],
    copula_list = copula_fit[["copula_list"]],
    n_cores = 2L,
    family_use = "nb",
    input_data = data_list[["covariate"]],
    new_covariate = data_list[["new_covariate"]],
    important_feature = copula_fit[["important_feature"]],
    filtered_gene = data_list[["filtered_gene"]],
    parallelization = "parallel"
    )

RNGkind("Mersenne-Twister")  # reset
```

### Step 6: create SingleCellExperiment object using simulated data

After simulating the data, we can create a `SingleCellExperiment` object
as follows.

```{r}
simu_sce <- SingleCellExperiment(list(counts = newcount_mat), 
                                 colData = data_list[["new_covariate"]])
names(assays(simu_sce)) <- "counts"

# save gene meta data
rowData(simu_sce) <- rowData(example_sce)
```

## Visualization

The simulated data can be visualized using a UMAP plot as follows.

```{r}
logcounts(example_sce) <- log1p(counts(example_sce))
logcounts(simu_sce) <- log1p(counts(simu_sce))
set.seed(123)
compare_figure <- scDesignPop::plotReducedDimPop(
    ref_sce = example_sce,
    sce_list = list(simu_sce), 
    name_vec = c("Reference", "scDesignPop"),
    assay_use = "logcounts", 
    if_plot = TRUE, 
    color_by = "cell_type", point_size = 1,
    n_pc = 30)
plot(compare_figure$p_umap)
```

## Create and visualize pseudobulk expression data

Next, we create pseudobulk data by normalizing the SCE object in the data using `log1p` and then aggregate using `mean`.  We specify the gene at the trans-eQTL loci as well using the `createPbulkExprGeno` function. Finally, we visualize the pseudobulk expression for the gene at this loci.

```{r,fig.height=3,fig.width=10}
res_list <- createPbulkExprGeno(sce_list = list("Reference" = example_sce,
                                                "Simulated" = simu_sce),
                                eqtlgeno = example_eqtlgeno,
                                feature_sel = "ENSG00000026025",
                                celltype_sel = "monoc",
                                eqtl_snp = "10:17010798",
                                normalize_type = "log1p",
                                aggregate_type = "mean",
                                slot_name = "logcounts",
                                overwrite = TRUE,
                                if_plot = TRUE)
res_list[["p_pbulk"]]
```

## Session information
```{r}
sessionInfo()
```
