---
title: "Model cell type proportions for new individuals"
author: 
  - name: Chris Dong
    affiliation:
    - Department of Statistics and Data Science, University of California, Los Angeles
    email: cycd@g.ucla.edu
  - name: Yihui Cen
    affiliation:
    - Department of Computational Medicine, University of California, Los Angeles
    email: yihuicen@g.ucla.edu
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('scDesignPop')`"
vignette: >
  %\VignetteIndexEntry{scDesignPop-celltypepropmodel-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{css, echo=FALSE}
pre {
  white-space: pre !important;
  overflow-x: scroll !important;
}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    warning = FALSE,
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
tools::R_user_dir("scDesignPop", which="cache")
```

## Introduction

scDesignPop provide modeling of the cell type proportions for generating new cell type proportions for new simulated individuals. The default covariates used are the PCs of the genotypes, which we will show in the following tutorial.

## Library and data preparation

Here, we load the `example_sce` data with both the genotype PCs of the training `example_sce` data and those of the new simulated individuals.

```{r}
load("/home/ycen/proj/scEQTLsim/data/tutorials/example_sce.rda")
load("/home/ycen/proj/scEQTLsim/data/tutorials/example_genoPCs.rda")
othercov_new <- dplyr::select(genopc_new, indiv)

head(genopc_train)
head(genopc_new)
```

## Celltype proportion modeling

We model the cell type proportions with a Multinomial model using genotype PCs as covariates. We also model the total cell number per individuals with a log-normal distribution. Here, `indiv_colname` is used to specify the shared column name for individual ids in both the `colData(example_sce)` and `genopc_train`. We use `cellstate_colname` to specify the column name for the cell states or cell types in `colData(example_sce)`.

```{r}
set.seed(123)
simu_cellprop_list <- scDesignPop::simuCellProportion(
  sce = example_sce,
  genoPC = genopc_train,
  new_genoPC = genopc_new,
  new_othercov = othercov_new,
  PCnum = 30L,
  cov_colnames = NULL,
  indiv_colname = "indiv",
  cellstate_colname = "cell_type",
  cn_model_family = "lognormal",    # cell number model
  cp_model_family = "MN",  # cell proportion model
  cp_intercept = TRUE
)
```

## Showing the covariates for new individuals

The covariates including the cell types for the new simulated individuals will be contained in the following data frame, which can be given to the `new_covariate` parameter in function `constructDataPop()`.

```{r}
head(simu_cellprop_list[["simu_cov"]])
```

## Visualizing the cell type proportion structures

By specifying the colors for each cell type, we can visualize the cell type proportion structure between the original data and simulated data after ordering individuals with the cell number of the first cell type. If no colors are specified for cell types, the function will still plot the data using the R default colors and ordering individuals based on the cell number of the first cell type after factorizing the cell types of the given data. 

```{r}
color_vec <- c(
  "cd4nc"   = "#5E3C99",
  "bmem"    = "#FFEDA0",
  "monoc"   = "#FD8D3C",
  "mononc"  = "#FC4E2A"
)


p1 <- scDesignPop::plotCellProp(colData(example_sce), 
                   title = "Original 40 individuals",
                   color_vec = color_vec, 
                   cellstate_colname = "cell_type",
                   indiv_colname = "indiv",
                   width = 1, linewidth = 0.02)

p2 <- scDesignPop::plotCellProp(simu_cellprop_list[["simu_cov"]], 
                   title = "Simulated 982 individuals",
                   color_vec = color_vec, 
                   cellstate_colname = "cell_type",
                   indiv_colname = "indiv",
                   width = 1, linewidth = 0.01)


cowplot::plot_grid(p1, p2)
```


