---
title: "Modify eQTL effect for eGenes / non-eGenes"
author: 
  - name: Chris Dong
    affiliation:
    - Department of Statistics and Data Science, University of California, Los Angeles
    email: cycd@g.ucla.edu
  - name: Yihui Cen
    affiliation:
    - Department of Computational Medicine, University of California, Los Angeles
    email: yihuicen@g.ucla.edu
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('scDesignPop')`"
vignette: >
  %\VignetteIndexEntry{scDesignPop-modifyEQTL-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---
```{css, echo=FALSE}
pre {
  white-space: pre !important;
  overflow-x: scroll !important;
}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
tools::R_user_dir("scDesignPop", which="cache")
```

## Introduction

scDesignPop enables users to modify the estimated cell-type-specific eQTL effect sizes and generate realistic scRNA-seq data with altered ground truths. Such data allow downstream tasks such as systematic benchmarking various eQTL mapping methods with different ground truths.

## Load libraries and data

Here, we use an example SingleCellExperiment object `example_sce` with $982$ genes and $7998$ cells and an example eQTL genotype dataframe `example_eqtlgeno` to demonstrate the tutorial. These two objects contains the gene expression and SNP genotypes of $40$ anonymized individuals while the eQTL genotype dataframe provides $2792$ putative cell-type-specific eQTLs.

```{r, message=FALSE, warning=FALSE, results='hide'}
library(scDesignPop)
library(SingleCellExperiment)
library(SummarizedExperiment)
library(ggplot2)

data("example_sce")
data("example_eqtlgeno")
```

## Model fitting

### Step 1: construct a data list

To run scDesignPop, a list of data is required as input. This is done using the `constructDataPop` function. A `SingleCellExperiment` object and an `eqtlgeno` dataframe are the two main inputs needed. The `eqtlgeno` dataframe consists of eQTL annotations (it must have cell state, gene, SNP, chromosome, and position columns at a minimum), and genotypes across individuals (columns) for every SNP (rows). The structure of an example `eqtlgeno` dataframe is given below.

```{r}
data_list <- constructDataPop(
    sce = example_sce,
    eqtlgeno_df = example_eqtlgeno,
    new_covariate = as.data.frame(colData(example_sce)),
    overlap_features = NULL,
    sampid_vec = NULL,
    copula_variable = "cell_type",
    slot_name = "counts",
    snp_mode = "single",
    celltype_colname = "cell_type",
    feature_colname = "gene_id",
    snp_colname = "snp_id",
    loc_colname = "POS",
    chrom_colname = "CHR",
    indiv_colname = "indiv",
    prune_thres = 0.9
    )
```

### Step 2: fit marginal model

Next, a marginal model is specified to fit each gene using the `fitMarginalPop` function.  
Here we use a Negative Binominal mixed model using `"nb"` option. We specify `"(1|indiv) + cell_type"` for the mean formula, where `(1|indiv)` specifies the random intercept using `indiv` as variable.

```{r}
marginal_list <- fitMarginalPop(
    data_list = data_list,
    mean_formula = "(1|indiv) + cell_type",
    model_family = "nb",
    interact_colnames = "cell_type",
    parallelization = "parallel",
    n_threads = 20L,
    loc_colname = "POS",
    snp_colname = "snp_id",
    celltype_colname = "cell_type",
    indiv_colname = "indiv",
    filter_snps = TRUE,
    snpvar_thres = 0,
    force_formula = FALSE,
    data_maxsize = 1
    )
```

### Step 3: fit a Gaussian copula

The third step is to fit a Gaussian copula using the `fitCopulaPop` function.

```{r}
set.seed(123, kind = "L'Ecuyer-CMRG")

copula_fit <- fitCopulaPop(
    sce = example_sce,
    assay_use = "counts",
    input_data = data_list[["new_covariate"]],
    marginal_list = marginal_list,
    family_use = "nb",
    copula = "gaussian",
    n_cores = 2L,
    parallelization = "parallel"
    )

RNGkind("Mersenne-Twister")  # reset
```

### Step 4. modify cell-typse-specific eQTL effect

To define a cell-type-specific eQTL gene (eGene), we modify modify the fitted parameters using the `modifyMarginalModels` function. Here, we increase the cell-type-specific eQTL effect size for `ENSG00000002549` gene by a $\log2$-fold-change relative to the fitted value using `eqtl_log2fc = 3` in `mononc` cell type.

```{r}
marginal_mod <- modifyMarginalModels(
    marginal_list = marginal_list,
    eqtlgeno_list = data_list[["eqtl_geno_list"]],
    features = "ENSG00000002549",  # "ENSG00000163221",
    celltype = "mononc",  # "cd4nc",
    eqtl_log2fc = 3,
    mean_baseline = NULL,
    eqtl_baseline = NULL,
    mean_baseline_only = FALSE,
    eqtl_baseline_only = FALSE,
    disp_scaling = FALSE,
    mod_scale = "link",
    verbose = TRUE
    )
```

### Step 5: extract parameters

The fifth step is to compute the mean, sigma, and zero probability parameters using the `extractParaPop` function.

```{r}
para_new <- extractParaPop(
    sce = example_sce,
    assay_use = "counts",
    marginal_list = marginal_mod[["marginal_list"]],
    n_cores = 2L,
    family_use = "nb",
    indiv_colname = "indiv",
    new_covariate = data_list[["new_covariate"]],
    new_eqtl_geno_list = data_list[["eqtl_geno_list"]],
    data = data_list[["covariate"]],
    parallelization = "parallel"
    )
```

### Step 6: simulate counts

The sixth step is to simulate counts using the `simuNewPop` function.

```{r}
set.seed(123, kind = "L'Ecuyer-CMRG")

newcount_mat <- simuNewPop(
    sce = example_sce,
    mean_mat = para_new[["mean_mat"]],
    sigma_mat = para_new[["sigma_mat"]],
    zero_mat = para_new[["zero_mat"]],
    quantile_mat = NULL,
    copula_list = copula_fit[["copula_list"]],
    n_cores = 2L,
    family_use = "nb",
    nonnegative = TRUE,
    input_data = data_list[["covariate"]],
    new_covariate = data_list[["new_covariate"]],
    important_feature = copula_fit[["important_feature"]],
    filtered_gene = data_list[["filtered_gene"]],
    parallelization = "parallel"
    )

RNGkind("Mersenne-Twister")  # reset
```

### Step 7: create SingleCellExperiment object using simulated data

After simulating the data, we create a `SingleCellExperiment` object as follows.

```{r}
simu_sce <- SingleCellExperiment(list(counts = newcount_mat),
                                 colData = data_list[["new_covariate"]])
```


## Create pseudobulk expression data

To create pseudobulk data, we first define a named list with the SCE objects of interest. Here we name `example_sce` as "Reference", and the simulated SCE object with modified effect size as "Simulated".

```{r}
sce_list <- list("Reference" = example_sce,
                 "Simulated" = simu_sce)
```

Next, we normalize each SCE object in the data using `log1p` and then aggregate using `mean`. This can be done in one step using the `createPbulkExprGeno` function. 

We also specify the all cells (denoted as `"allct"`), and the cell type that we wish to visualize with `c("allct", "mononc")`. The SNP loci specified correspond to the cell types in the same order.

```{r}
res_list <- createPbulkExprGeno(sce_list = sce_list,
                                eqtlgeno = example_eqtlgeno,
                                feature_sel = "ENSG00000185201",
                                celltype_sel = c("allct", "mononc"),
                                eqtl_snp = c("11:320836", "11:293188"),
                                normalize_type = "log1p",
                                aggregate_type = "mean",
                                slot_name = "logcounts",
                                overwrite = TRUE,
                                if_plot = TRUE)
```

One of the objects, `res_df`, in the output list is a dataframe with each row containing the pseudobulk values corresponding to each individual and their genotypes, along with other important columns such as cell types, aggregation used, normalization used, etc.

```{r}
head(res_list[["res_df"]])
```

## Visualize pseudobulk expression

Next, we show the pseudobulk expression versus genotype using violin plots across the two SCE objects and the cell types using the ggplot object, `p_pbulk`, generated in the output list. Each violin plot has a boxplot within and the mean marked by a red dot.

```{r, fig.height= 3, fig.width=6}
res_list[["p_pbulk"]]
```

Alternatively, users may customize the plot using additional `ggplot2` commands or generate their own plot directly using the `res_df` dataframe. We leave the latter part out of this tutorial.

```{r, fig.height= 6, fig.width=6}
res_list[["p_pbulk"]] +
    theme_classic() +
    labs(title = "ENSG00000002549",
                  y = "Normalized expression",
                  x = "Genotype") +
    theme(
        plot.title = element_text(size = 14, hjust = 0,  # left title
                                  face = "bold"),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 12),
        strip.text = element_text(size = 12),  # facet font size
        aspect.ratio = 1,
        legend.position = "none"
    )
```

## Session information
```{r}
sessionInfo()
```
