---
title: "Power analysis based on a fitted scDesignPop marginal model"
author: 
  - name: Chris Dong
    affiliation:
    - Department of Statistics and Data Science, University of California, Los Angeles
    email: cycd@g.ucla.edu
  - name: Yihui Cen
    affiliation:
    - Department of Computational Medicine, University of California, Los Angeles
    email: yihuicen@g.ucla.edu
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('scDesignPop')`"
vignette: >
  %\VignetteIndexEntry{scDesignPop-poweranalysisD-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---
```{css, echo=FALSE}
pre {
  white-space: pre !important;
  overflow-x: scroll !important;
}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    message = FALSE,
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
tools::R_user_dir("scDesignPop", which="cache")
```

## Step 1: running power analysis

Given fitted marginal model, scDesignPop can perform simulation-based power analysis for a
specific gene-SNP pair across cell types using the `runPowerAnalysis`
function. Here, we choose the negative binomial mixed model using `"nb"`
option.

```{r}
library(scDesignPop)
library(SingleCellExperiment)
load("/home/ycen/proj/scEQTLsim/data/tutorials/marginal_list_sel.rda")
set.seed(123)
power_data <- runPowerAnalysis(marginal_list = marginal_list_sel,
                               marginal_model = "nb",
                               geneid = "ENSG00000163221",
                               snpid = "1:153337943",
                               type_specific = "cell_type",
                               type_vector = c("bmem", "monoc"),
                               methods = c("poisson","pseudoBulkLinear"),
                               nindivs = c(50, 200),
                               ncells = c(10),
                               alpha = 0.05,
                               power_nsim = 50,
                               snp_number = 10,
                               gene_number = 200,
                               CI_nsim = 1000,
                               CI_conf = 0.05,
                               ncores = 50L)
```

## Step 2: visualizing power results

The power analysis results can be visualized using the
`visualizePowerResult` function.

```{r}
visualizePowerResult(power_result = power_data,
                     celltypes = c("bmem", "monoc"),
                     x_axis = "nindiv",
                     y_axis = "ncell",
                     col_group = "method",
                     geneid = "ENSG00000163221",
                     snpid = "1:153337943")
```

